---
title: "Damage Analysis of Storms"
output:
  html_document:
    keep_md: yes
---
# Synopsis
This document presents an analysis of weather event types and their negative impacts on economic factors (property and crop damages) and personal health (number of fatalities and number of injuries). It is analyzed which weather events have the largest impacts on each, health and economy.

# Data Processing
```{r}
library(dplyr)
setwd ("c:/r_dat/repro_pa2")
storm <- read.csv(bzfile(paste("repdata_data_StormData.csv.bz2", sep = "")),sep=",")
#storm<-tbl_df(read.csv("storm_sample2.csv"))

# The EVTYPE levels are not clean.  The following section is an attempt to clean up
# and categorize all levels to the 47 listed in the Storm Data Documentation in Section 2.1.1
storm$EVTYPE <- gsub("^(HEAT).*", "HEAT", storm$EVTYPE)
storm$EVTYPE <- gsub("^(RECORD HEAT).*", "HEAT", storm$EVTYPE)
storm$EVTYPE <- gsub("^(EXTREME HEAT).*", "HEAT", storm$EVTYPE)
storm$EVTYPE <- gsub("^(Heat).*", "HEAT", storm$EVTYPE)
storm$EVTYPE <- gsub("^(EXCESSIVE HEAT).*", "HEAT", storm$EVTYPE)
storm$EVTYPE <- gsub("^(TSTM).*", "THUNDER STORM", storm$EVTYPE)
storm$EVTYPE <- gsub("^(THUNDERSTORM).*", "THUNDER STORM", storm$EVTYPE)
storm$EVTYPE <- gsub("^(TROPICAL STORM).*", "TROPICAL STORM", storm$EVTYPE)
storm$EVTYPE <- gsub("^(FLASH FLOOD).*", "FLOOD", storm$EVTYPE)
storm$EVTYPE <- gsub("^(WIND).*", "WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("^(STRONG WIND).*", "WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HIGH WIND).*", "WIND", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HURRICANE).*", "HURICCANE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(SNOW).*", "SNOW", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HEAVY SNOW).*", "SNOW", storm$EVTYPE)
storm$EVTYPE <- gsub("^(FIRE).*", "FIRE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(WILD/FOREST FIRE).*", "FIRE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(WILDFIRE).*", "FIRE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(WILD FIRES).*", "FIRE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HAIL).*", "HAIL", storm$EVTYPE)
storm$EVTYPE <- gsub("^(BLIZZARD).*", "BLIZZARD", storm$EVTYPE)
storm$EVTYPE <- gsub("^(COLD).*", "COLD", storm$EVTYPE)
storm$EVTYPE <- gsub("^(WINTER WEATHER).*", "COLD", storm$EVTYPE)
storm$EVTYPE <- gsub("^(EXTREME COLD).*", "COLD", storm$EVTYPE)
storm$EVTYPE <- gsub("^(RIP).*", "RIP", storm$EVTYPE)
storm$EVTYPE <- gsub("^(FOG).*", "FOG", storm$EVTYPE)
storm$EVTYPE <- gsub("^(DENSE FOG).*", "FOG", storm$EVTYPE)
storm$EVTYPE <- gsub("^(AVALANCHE).*", "AVALANCHE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(AVALANCE).*", "AVALANCHE", storm$EVTYPE)
storm$EVTYPE <- gsub("^(RAIN).*", "RAIN", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HEAVY RAIN).*", "RAIN", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HIGH SURF).*", "SURF", storm$EVTYPE)
storm$EVTYPE <- gsub("^(HEAVY SURF).*", "SURF", storm$EVTYPE)
storm$EVTYPE <- gsub("^(SURF).*", "SURF", storm$EVTYPE)
```
## Data preparation for the results of question 1
We group the storm data by EVTYPE and calculate the sum of all fatalities and injuries by each event type for the grouped stormdata with the summarize function. After that, we filter all events in which fatalities OR injuries occured. Then we sort the data by the sum of fatalities and then by the sum of injuries.
```{r}
by_events<-group_by(storm,EVTYPE)
person_damage<-summarize(by_events,sum_fat=sum(FATALITIES),sum_inj=sum(INJURIES))
pdamages<-filter(person_damage,sum_inj>0 | sum_fat>0)
pdamages<-arrange(pdamages, desc(sum_fat), desc(sum_inj))
```
## Data preparation for the results of question 2
For the property damage values and the crop damage values, the units of measurements (PROPDMGEXP, CROPDMGEXP) are replaced by numeric values in the storm data set. After that the values in PROPDMG and CROPDMG are multiplied by the numeric values of the units of measurement. We then group the processed data by the weather event variable EVTYPE. After that the amounts of damages for both types (property and crop) are calculated (sum_precondam and sum_crecondam). The data are filtered, sorted in the last step.
```{r}
storm$PROPDMGEXP <- gsub("\\-|\\+|\\?|0", "1", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("h|H|2","100", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("k|K|3","1000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("4","10000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("5","100000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("m|M|6","1000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("7","10000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("8","100000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("b|B|9","1000000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- as.numeric(storm$PROPDMGEXP)

storm$CROPDMGEXP <- gsub("\\-|\\+|\\?|0", "1", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("h|H|2","100", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("k|K|3","1000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("4","10000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("5","100000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("m|M|6","1000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("7","10000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("8","100000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("b|B|9","1000000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- as.numeric(storm$CROPDMGEXP)

storm$PRECONDAM <- storm$PROPDMG * storm$PROPDMGEXP
storm$CRECONDAM <- storm$CROPDMG * storm$CROPDMGEXP
storm$ECONDAM <- storm$PRECONDAM + storm$CRECONDAM

by_events_econ<-group_by(storm,EVTYPE)
econ_damage<-summarize(by_events_econ, sum_econdam=sum(ECONDAM))
edamages<-filter(econ_damage,sum_econdam>0)
edamages<-arrange(edamages, desc(sum_econdam))
```

# Results
## Results Q1
```{r}
pdamages<-head(pdamages,10)
print(pdamages)
par(mfrow=c(1,2))
barplot(pdamages$sum_fat,names.arg=pdamages$EVTYPE,cex.names=0.5,las=2)
barplot(pdamages$sum_inj,names.arg=pdamages$EVTYPE,cex.names=0.5,las=2)
```
  
## Results Q2
```{r}
edamages<-head(edamages,10)
print(edamages)
barplot(edamages$sum_econdam, names.arg=edamages$EVTYPE,cex.names=0.5,las=2)
```