---
title: "Damage Analysis of Storms"
output:
  html_document:
    keep_md: yes
---
# Synopsis
This document presents an analysis of weather event types and their negative impacts on economic factors (property and crop damages) and personal health (number of fatalities and number of injuries). It is analyzed which weather events have the largest impacts on each, health and economy.

# Data Processing
```{r}
library(plyr)
library(dplyr)
setwd ("c:/r_dat/repro_pa2")
storm <- read.csv(bzfile(paste("repdata_data_StormData.csv.bz2", sep = "")),sep=",")
#storm<-tbl_df(read.csv("storm_sample.csv"))

```
## Data preparation for the results of question 1
We group the storm data by EVTYPE and calculate the sum of all fatalities and injuries by each event type for the grouped stormdata with the summarize function. After that, we filter all events in which fatalities OR injuries occured. Then we sort the data by the sum of fatalities and then by the sum of injuries.
```{r}
by_events<-group_by(storm,EVTYPE)
person_damage<-summarize(by_events,sum_fat=sum(FATALITIES),sum_inj=sum(INJURIES))
pdamages<-filter(person_damage,sum_inj>0 | sum_fat>0)
```
## Data preparation for the results of question 2
For the property damage values and the crop damage values, the units of measurements (PROPDMGEXP, CROPDMGEXP) are replaced by numeric values in the storm data set. After that the values in PROPDMG and CROPDMG are multiplied by the numeric values of the units of measurement. We then group the processed data by the weather event variable EVTYPE. After that the amounts of damages for both types (property and crop) are calculated (sum_precondam and sum_crecondam). The data are filtered, sorted in the last step.
```{r}
storm$PROPDMGEXP <- gsub("\\-|\\+|\\?|0", "1", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("h|H|2","100", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("k|K|3","1000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("4","10000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("5","100000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("m|M|6","1000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("7","10000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("8","100000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- gsub("b|B|9","1000000000", storm$PROPDMGEXP)
storm$PROPDMGEXP <- as.numeric(storm$PROPDMGEXP)

storm$CROPDMGEXP <- gsub("\\-|\\+|\\?|0", "1", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("h|H|2","100", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("k|K|3","1000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("4","10000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("5","100000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("m|M|6","1000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("7","10000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("8","100000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- gsub("b|B|9","1000000000", storm$CROPDMGEXP)
storm$CROPDMGEXP <- as.numeric(storm$CROPDMGEXP)

storm$PRECONDAM <- storm$PROPDMG * storm$PROPDMGEXP
storm$CRECONDAM <- storm$CROPDMG * storm$CROPDMGEXP
storm$ECONDAM <- storm$PRECONDAM + storm$CRECONDAM

by_events_econ<-group_by(storm,EVTYPE)
econ_damage<-summarize(by_events_econ, sum_econdam=sum(ECONDAM))
edamages<-filter(econ_damage,sum_econdam>0)
edamages<-arrange(edamages, desc(sum_econdam))
edamages$sum_econdam<-edamages$sum_econdam/1000000000
```

# Results
## Results Q1
In order to analyse Q1, we first print the TOP 10 harmful events by their number of fatalities and injuries! After that we present a panel bar plot to illustrate the results graphically.
```{r}
pdamages_fat<-head(arrange(pdamages, desc(sum_fat)),10)
pdamages_inj<-head(arrange(pdamages, desc(sum_inj)),10)

print(pdamages_fat)
print(pdamages_inj)

par(mfrow=c(1,2))
barplot(pdamages_fat$sum_fat,names.arg=pdamages_fat$EVTYPE,cex.names=0.5,las=2)
barplot(pdamages_inj$sum_inj,names.arg=pdamages_inj$EVTYPE,cex.names=0.5,las=2)
```


## Results Q2
In order to analyse Q2, we first print the TOP 10 harmful events by their ecnomomic impact! After that we present a bar plot to illustrate the results graphically.
```{r}
edamages<-head(edamages,20)
print(edamages)
barplot(edamages$sum_econdam, names.arg=edamages$EVTYPE,cex.names=0.5,las=2)
```